---
title: "Replicating a Published Butterfly Biodiversity Analysis in R"
subtitle: "ANOVA, detrended correspondence analysis (DCA), and correspondence analysis (CA)"
author: "Nikolay Winslow"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Overview

This report replicates the main analyses from Loos et al. (2014), a study
of butterfly communities across different land-use types. The goals are
to:

1.  Prepare the data so that only arable and grassland sites remain.
2.  Reproduce the study's descriptive statistics and ANOVA results.
3.  Reproduce the detrended correspondence analysis (DCA) shown in
    Figure 2 of the paper (without the environmental vectors).
4.  Run a correspondence analysis (CA) on transformed data and compare it
    with the DCA.

The data is publicly available at
[Dryad](https://doi.org/10.5061/dryad.97s1k). Where my results differ
from the published ones, the differences are noted and discussed.

## Setup

```{r}
library(readxl)
library(vegan)
library(dplyr)

# Dataset: Loos et al. (2014) butterfly survey, publicly available on Dryad
# https://doi.org/10.5061/dryad.97s1k
data <- read_excel("Loos+et+al+2014+Butterflies.xls")
```

## 1. Data Preparation

The dataset contains site identifiers, species counts, and a final block
of environmental variables. The steps below:

-   remove forest sites (`LUtype == "F"`) so only arable and grassland
    sites remain,
-   split the species counts from the environmental variables, and
-   drop species that never occurred in the remaining sites (some
    species were recorded only in forest).

```{r}
# Keep only arable and grassland sites
data2 <- data %>%
  filter(LUtype != "F")

# Column layout: column 1 = site ID, then species counts, then the final
# 16 columns are environmental variables
n_env <- 16

env     <- data2[, (ncol(data2) - n_env + 1):ncol(data2)]
species <- data2[, 2:(ncol(data2) - n_env)]

# Remove species with no occurrences in the remaining sites
species <- species[, colSums(species) > 0]

# Land-use type as a factor, reused for colouring the ordination plots
lu      <- factor(data2$LUtype)
lu_cols <- seq_along(levels(lu))

cat("Sites:", nrow(species), "| Species:", ncol(species), "\n")
```

## 2. Descriptive Statistics and ANOVA

```{r}
# Totals and ranges
total_individuals <- sum(species)
num_species       <- ncol(species)
richness          <- rowSums(species > 0)  # species richness per site
abundance         <- rowSums(species)      # individuals per site

summary_stats <- list(
  total_individuals = total_individuals,
  num_species       = num_species,
  richness_range    = range(richness),
  abundance_range   = range(abundance)
)
print(summary_stats)

# Most abundant species (proportion of all individuals)
species_rel <- sort(colSums(species) / total_individuals, decreasing = TRUE)
print(head(species_rel, 12))
```

```{r}
# Does species richness differ between land-use types?
anova_richness <- aov(richness ~ LUtype, data = data2)
summary(anova_richness)

# Does abundance differ between land-use types?
anova_abundance <- aov(abundance ~ LUtype, data = data2)
summary(anova_abundance)
```

**Comparison with the paper.** The statistics almost entirely match the
published values. The exceptions are the range of species richness per
site (3-47 here versus 3-45 in the paper) and the total counts of
individuals and species, which differ slightly, possibly because of
known taxonomic inconsistencies in the dataset. The most abundant
species and the ANOVA results for richness and abundance match the
paper. As the paper reports, the ANOVA results show little difference
between arable and grassland sites in either richness or abundance.

## 3. Detrended Correspondence Analysis (DCA)

```{r}
#| fig-cap: "DCA of butterfly assemblages by site, coloured by land-use type (replication of Figure 2 without environmental vectors)."
dca_model <- decorana(species)

plot(dca_model, type = "n", xlim = c(-1.5, 2), ylim = c(-1, 1.5))
points(dca_model, display = "sites", col = as.integer(lu), pch = 19)
legend("topright", legend = levels(lu), col = lu_cols, pch = 19)
```

**Comparison with the paper.** The DCA was carried out successfully and
matches Figure 2 of the study.

## 4. Correspondence Analysis (CA)

Butterfly counts are highly skewed, so the counts were log-transformed
before running CA, using the `log` method in `vegan::decostand()`
(log(x) + 1 for non-zero counts; zeros stay zero).

```{r}
species_log <- decostand(species, method = "log")

ca_model <- cca(species_log)

# Variance explained by each axis
eig_vals      <- eigenvals(ca_model)
var_explained <- eig_vals / sum(eig_vals)

cat(sprintf("Variance explained by CA1: %.1f%%\n", 100 * var_explained[1]))
cat(sprintf("Variance explained by CA2: %.1f%%\n", 100 * var_explained[2]))
```

```{r}
#| fig-cap: "CA of log-transformed butterfly data by site, coloured by land-use type."
plot(ca_model, type = "n")
points(ca_model, display = "sites", col = as.integer(lu), pch = 19)
legend("topright", legend = levels(lu), col = lu_cols, pch = 19)
```

**Interpretation.** The first two CA axes together explain only about
10% of the variance, which indicates that no single dominant factor
explains the distribution of the butterflies. An arch effect is visible
to some extent in the CA: points condense near the origin, while values
spread farther apart on CA1 as CA2 decreases. This differs from the
DCA, where arable sites are spread along DCA1 and grassland sites along
DCA2, forming a very loose L-shaped pattern. The parabolic structure of
the CA makes the groupings look more discrete than they are, whereas the
detrended analysis shows only a vague trend. Given the low variance
explained by each CA axis and CA's tendency to produce an arch, the DCA
appears to give the more accurate representation of the data.

## Conclusions

-   The descriptive statistics, most abundant species, and ANOVA results
    replicate the published findings, apart from small differences in the
    richness range and in total individual and species counts.
-   Species richness and abundance differ little between arable and
    grassland sites.
-   The DCA reproduces Figure 2 of the paper.
-   A CA of log-transformed data explains about 10% of the variance on
    its first two axes and shows an arch effect, so the DCA is the better
    representation of these data.

## Reproducibility

```{r}
sessionInfo()
```

## Reference

Loos, J., Dorresteijn, I., Hanspach, J., Fust, P., Rakosy, L., &
Fischer, J. (2014). Low-intensity agricultural landscapes in Transylvania
support high butterfly diversity: implications for conservation. *PLoS
ONE, 9*(7), e103256.

Loos, J., Dorresteijn, I., Hanspach, J., Fust, P., Rakosy, L., &
Fischer, J. (2015). *Data from: Low-intensity agricultural landscapes in
Transylvania support high butterfly diversity: implications for
conservation* [Dataset]. Dryad. <https://doi.org/10.5061/dryad.97s1k>
